Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation

### 2022-01-18 ms/SI.Rmd
Update MS

**`r s_figure("grain-size")`**. Effect of grain size on sampling redundancy. (a), Percent of grid cells with redundancy > 0.5 by grain size (side length of square grid cells set to 10 km, 20 km, 30 km, or 40 km). (b), Percent of grid cells with redundancy = 0 by grain size. (c), Distribution of redundancy values by grain size.

@@

**`r s_figure("envir")`**. Reproductive mode and environmental data in 20 km × 20 km grid cells across Japan. (a), rolling mean area (km^2^) in 20 km latitudinal bands.  (b), % apomictic taxa. (c), annual precipitation. (d), percipitation seasonality (coefficient of variation). (e), mean annual temperature. (f), temperature seasonality (standard deviation × 100). Data excluded from models not shown in this and subsequent plots unless otherwise mentioned.

@@

**`r s_figure("abundance")`**. Observed number of specimens (a) and sampling redundancy (b) per 20 km grid cell in the ferns of Japan.

@@

**`r s_figure("raw-lat")`**. (a) Raw taxonomic diversity (b), phylogenetic diversity, and (c) functional diversity of the ferns of Japan plotted by latitude. For (b) and (c), raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths. Trendlines fit with general additive model using `r func("geom_smooth")` function in the `r pack("ggplot2")` R package [@Wickham2016].

@@

**`r s_figure("raw-biplot")`**. Relationships between observed functional and phylogenetic diversity and taxonomic richness in the ferns of Japan. (a), Phylogenetic diversity (PD). (b), Relative phylogenetic diversity (RPD). (c), Functional diversity (FD). (d), Relative functional diversity (RFD). Trendlines in (a, c) fit with general additive model using `r func("geom_smooth")` function in the `r pack("ggplot2")` R package [@Wickham2016].

@@

**`r s_figure("k-plot")`**. Selection of *k* for bioregions. (a), taxonomic bioregions. (b), phylogenetic bioregions. Color indicates optimal *k* selected selected by the `r func("optimal_phyloregion")` function in the R package `r pack("phyloregion")` (black, non-optimal; red, optimal).

@@

**`r s_figure("dendrogram")`**. Bioregion dendrograms. Tips are sites (names not shown). Dendrograms generated by clustering taxonomic (Sørensen) or phylogenetic (PhyloSor) distances between sites (see Materials and Methods). Color shows bioregion; bioregions not consisting of > `r english2(bioregion_cutoff)` grid-cells each are lumped into the "Other" category. (a), taxonomic bioregions. (b), phylogenetic bioregions. Dendrograms visualized with the `r pack("ggtree")` R package [@Yu2017].
\clearpage

@@

**`r s_figure("modelfit-not-ult")`**. Relationship between (a) phylogenetic diversity (PD) and (b) relative PD and % apomictic taxa, models inferred using phylogeny with branchlengths in units of expected genetic change (untransformed ML tree). Ribbon shows 95% confidence interval of model. Line fit with the focal predictor variable while averaging over other predictors. For RPD, the standard effect size (SES) was calculated by comparing raw values to a null distribution of 999 random values, and significance (*P*-value) calculated using a two-tailed test (see Materials and Methods).

@@

**`r s_figure("bioregions-env")`**. Scatterplots of grid cell membership in (a) taxonomic or (b) phylogenetic bioregions arranged by mean annual temperature and annual precipitation. Points shown with partial transparency; darker areas indicate overlapping points. Bioregions not consisting of > `r english2(bioregion_cutoff)` grid-cells each are lumped into the "Other" category.

>>>>

**`r s_figure("grain-size")`**. Effect of grain size on sampling redundancy. (A), Percent of grid cells with redundancy > 0.5 by grain size (side length of square grid cells set to 10 km, 20 km, 30 km, or 40 km). (B), Percent of grid cells with redundancy = 0 by grain size. (C), Distribution of redundancy values by grain size.

@@

**`r s_figure("envir")`**. Reproductive mode and environmental data in 20 km × 20 km grid cells across Japan. (A), rolling mean area (km^2^) in 20 km latitudinal bands.  (B), % apomictic taxa. (C), annual precipitation. (D), percipitation seasonality (coefficient of variation). (E), mean annual temperature. (F), temperature seasonality (standard deviation × 100). Data excluded from models not shown in this and subsequent plots unless otherwise mentioned.

@@

**`r s_figure("abundance")`**. Observed number of specimens (A) and sampling redundancy (B) per 20 km grid cell in the ferns of Japan.

@@

**`r s_figure("raw-lat")`**. (A) Raw taxonomic diversity (B), phylogenetic diversity, and (C) functional diversity of the ferns of Japan plotted by latitude. For (B) and (C), raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths. Trendlines fit with general additive model using `r func("geom_smooth")` function in the `r pack("ggplot2")` R package [@Wickham2016].

@@

**`r s_figure("raw-biplot")`**. Relationships between observed functional and phylogenetic diversity and taxonomic richness in the ferns of Japan. (A), Phylogenetic diversity (PD). (B), Relative phylogenetic diversity (RPD). (C), Functional diversity (FD). (D), Relative functional diversity (RFD). Trendlines in (a, c) fit with general additive model using `r func("geom_smooth")` function in the `r pack("ggplot2")` R package [@Wickham2016].

@@

**`r s_figure("k-plot")`**. Selection of *k* for bioregions. (A), taxonomic bioregions. (B), phylogenetic bioregions. Color indicates optimal *k* selected selected by the `r func("optimal_phyloregion")` function in the R package `r pack("phyloregion")` (black, non-optimal; red, optimal).

@@

**`r s_figure("dendrogram")`**. Bioregion dendrograms. Tips are sites (names not shown). Dendrograms generated by clustering taxonomic (Sørensen) or phylogenetic (PhyloSor) distances between sites (see Materials and Methods). Color shows bioregion; bioregions not consisting of > `r english2(bioregion_cutoff)` grid-cells each are lumped into the "Other" category. (A), taxonomic bioregions. (B), phylogenetic bioregions. Dendrograms visualized with the `r pack("ggtree")` R package [@Yu2017].
\clearpage

@@

**`r s_figure("modelfit-not-ult")`**. Relationship between (A) phylogenetic diversity (PD) and (B) relative PD and % apomictic taxa, models inferred using phylogeny with branchlengths in units of expected genetic change (untransformed ML tree). Ribbon shows 95% confidence interval of model. Line fit with the focal predictor variable while averaging over other predictors. For RPD, the standard effect size (SES) was calculated by comparing raw values to a null distribution of 999 random values, and significance (*P*-value) calculated using a two-tailed test (see Materials and Methods).

@@

**`r s_figure("bioregions-env")`**. Scatterplots of grid cell membership in (A) taxonomic or (B) phylogenetic bioregions arranged by mean annual temperature and annual precipitation. Points shown with partial transparency; darker areas indicate overlapping points. Bioregions not consisting of > `r english2(bioregion_cutoff)` grid-cells each are lumped into the "Other" category.

### 2022-01-18 ms/manuscript.Rmd
Update MS

One particularly valuable characteristic of the Japanese fern flora is the availability of data on reproductive mode (known for `r repro_count$known$total` of `r repro_count$known$total + repro_count$unknown$total` native pteridophyte taxa excluding hybrids [`r repro_count$known$percent`]) [@Ebihara2019b].
Reproductive mode is likely to affect population-level genetic diversity [@Bengtsson2003], and thereby higher-level biodiversity [@Krueger-Hadfield2019].
In ferns, clonally reproducing apomictic species are often polyploid hybrids that share identical plastid genotypes among taxa and within populations [@Grusz2009; @Chao2012a; @Hori2014; @Hori2019].
Therefore, we expect high prevalence of apomictic species within a community to decrease phylogenetic diversity.

Japan is situated along a latitudinal gradient spanning seasonal, temperate areas in the north to subtropical, mostly aseasonal islands in the south.
Furthermore, it is a mountainous country with great variation in elevation on the larger islands.
Plant distributions often reflect physiological adaptations to climate [e.g., precipitation and temperature/; @Woodward1987].
Therefore, we expect the spatial distribution of biodiversity to be determined by climatic variation in addition to reproductive mode.

@@

We measured phylogenetic diversity (PD) using the Faith's PD [the total branch length connecting all terminal taxa in a community, including the root of the phylogeny; @Faith1992].
To detect areas with unusually long or short branches, we calculated relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been transformed to equal length [@Mishler2014].

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native, non-hybrid ferns [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families.
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a\; `r s_figure("tree")`).
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016].
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

Water limitation is thought to play an important role in promoting apomixis in ferns especially in deserts [@Grusz2021], since ferns depend on water for transfer of sperm to egg, and apomictic plants would be able to reproduce in the absence of water.
A similar argument has been made to assert that areas with seasonal monsoons are linked to increased rates of apomictic ferns [@Liu2012; @Tanaka2014; @Picard2021], but in Japan, although the areas including high percentages of apomictic taxa are monsoonal, the degree of seasonal water limitation does not approach that of a desert.
The monsoon hypothesis is also not supported by our results, which show precipitation seasonality did not play a strong role in any model and was not correlated with % apomictic taxa.

@@

We found that grid-cells with significantly high amounts of biodiversity are better protected than the baseline protection rate at the medium protection level across all biodiversity metrics, and better than the baseline rate at the high protection level for PE (`r figure("conserv-status")`A).
The high protection status of grid-cells with high PE may be due to an emphasis on conserving endemic plant and animal species in Japanese southern islands such as Iriomote and Amami [@kankyosho2021b], which include many grid-cells with high PE (`r s_figure("protected-map")`D).
One possible reason for the generally high level of protection overall is that in the past, some ferns have been included in lists of rare and endemic plants for consideration in designation of conservation areas [A. Ebihara, pers. comm.; @kankyosho2021b].
Also, (as suggested previously) ferns may serve as bioindicators reflecting overall levels of biodiversity, particularly for other plant groups.

@@

Our analysis shows that the threat to fern biodiversity posed by Japanese deer differs strongly by biodiversity type; although grid-cells with significantly high FD, PD, and PE tend to have lower rates of deer occupancy compared to the baseline rate, grid-cells with high richness experience much greater co-occurrence with deer than the baseline rate (`r figure("conserv-status")`B).
This pattern is due to the fact that Japanese deer do not occur in the southern islands, which house high amounts of PD and PE; rather, areas with high species richness are primarily located in southern Honshu, Shikoku, and Kyushu, where deer expansion has been intense (`r s_figure("protected-map")`A).
This result highlights the importance of using multiple metrics of biodiversity for establishing conservation priorities.

@@

One element of biodiversity that we did not consider here is hybrids.
The Japanese fern flora includes `r n_hybrid_fern_taxa` hybrid taxa.
Although hybrids are often thought of as evolutionary "dead-ends", in some cases they are able to interbreed with diploids and thus contribute to diversification [@Barrington1989].
Future studies should consider the distribution of hybrids in Japan and their possible effects on biodiversity.
Methods for calculating PD would need to be modified to take into account multiple lineages contributing to one terminal taxon.

Finally, a long under-appreciated aspect of fern ecology is the role of the gametophyte.
Unlike seed plants, ferns have gametophytes that are capable of growing independently from the sporophyte.
Futhermore, the two stages of the life cycle have vastly different morphology and ecophysiology, and may even occur over partially to completely disjunct ranges [reviewed in @Pinson2016a].
Here, as in the vast majority of fern ecology studies, we consider only the sporophyte stage.
However, recent studies in Japan [@Ebihara2013; @Ebihara2019a] and elsewhere [@Nitta2017] have revealed different patterns in the distribution of gametophytes and sporophytes.
The comprehensive molecular sampling of sporophytes in Japan makes this area ideal for conducting high-throughput DNA sequencing analyses to compare patterns of biodiversity between life stages in ferns across large spatial scales.

@@

The authors thank Members of the Iwasaki Lab at the University of Tokyo for providing comments that improved the manuscript.
Shawn Laffan provided advice about CANAPE.
This study was supported by JSPS KAKENHI Grant Number 16H06279.

@@

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_spatial_phy.
A Docker image to run the code is available at https://hub.docker.com/r/joelnitta/japan_ferns_spatial_phy.
Data files needed to run the analysis and selected results files are available from Figshare at http://doi.org/10.6084/m9.figshare.16655263.
**Note to reviewers**: the Figshare DOI will not be accessible until the paper is accepted. During review, the data can be accessed from this link: https://figshare.com/s/de3d5a70e118208e56c5.

>>>>

One particularly valuable characteristic of the Japanese fern flora is the availability of data on reproductive mode, which is known for `r repro_count$known$total` of `r repro_count$known$total + repro_count$unknown$total` native pteridophyte taxa excluding hybrids [`r repro_count$known$percent`\; @Ebihara2019b].
Reproductive mode is likely to affect population-level genetic diversity [@Bengtsson2003], and thereby higher-level biodiversity [@Krueger-Hadfield2019].
In ferns, clonally reproducing apomictic species are often polyploid hybrids that share identical plastid genotypes among taxa and within populations [@Grusz2009; @Chao2012a; @Hori2014; @Hori2019].
Therefore, we expect high prevalence of apomictic species within a community to decrease phylogenetic diversity.

Japan is situated along a latitudinal gradient spanning seasonal, temperate areas in the north to subtropical, mostly aseasonal islands in the south.
Furthermore, it is a mountainous country with great variation in elevation on the larger islands.
Plant distributions often reflect physiological adaptations to climate [e.g., precipitation and temperature\; @Woodward1987].
Therefore, we expect the spatial distribution of biodiversity to be determined by climatic variation in addition to reproductive mode.

@@

We measured phylogenetic diversity (PD) using the Faith's PD [the total branch length connecting all terminal taxa in a community, including the root of the phylogeny\; @Faith1992].
To detect areas with unusually long or short branches, we calculated relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been transformed to equal length [@Mishler2014].

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native, non-hybrid ferns [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families.
The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a\; `r s_figure("tree")`].
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016].
The age of the fern clade is estimated to be `r phylogeny_age` ma.

@@

Water limitation is thought to play an important role in promoting apomixis in ferns especially in deserts [@Grusz2021], since ferns depend on water for transfer of sperm to egg, and apomictic plants would be able to reproduce in the absence of water.
A similar argument has been made to assert that areas with seasonal monsoons are linked to increased rates of apomixis in ferns [@Liu2012; @Tanaka2014; @Picard2021], but in Japan, although the areas including high percentages of apomictic taxa are monsoonal, the degree of seasonal water limitation does not approach that of a desert.
The monsoon hypothesis is not supported by our results, which did not show a particularly strong effect of precipitation seasonality in most models `r figure("coeff")`, nor a correlation with % apomictic taxa (`r s_table("corr")`).

@@

We found that grid-cells with significantly high amounts of biodiversity are similarly or better protected compared to the baseline protection rate at the medium protection level across all biodiversity metrics, and better than the baseline rate at the high protection level for PE (`r figure("conserv-status")`A).
The high protection status of grid-cells with high PE may be due to an emphasis on conserving endemic plant and animal species in Japanese southern islands such as Iriomote and Amami [@kankyosho2021b], which include many grid-cells with high PE (`r s_figure("protected-map")`D).
One possible reason for the generally high level of protection overall is that in the past, some ferns have been included in lists of rare and endemic plants for consideration in designation of conservation areas [A. Ebihara, pers. comm.\; @kankyosho2021b].
Also, (as suggested previously) ferns may serve as bioindicators reflecting overall levels of biodiversity, particularly for other plant groups.

@@

Our analysis shows that the threat to fern biodiversity posed by Japanese deer differs strongly by biodiversity type; although grid-cells with significantly high FD, PD, and PE tend to have lower rates of deer occupancy compared to the baseline rate in recent surveys or modeled data, grid-cells with high richness experience much greater co-occurrence with deer than the baseline rate (`r figure("conserv-status")`B).
This pattern is due to the fact that Japanese deer do not occur in the southern islands, which house high amounts of PD and PE; rather, areas with high species richness are primarily located in southern Honshu, Shikoku, and Kyushu, where deer expansion has been intense (`r s_figure("protected-map")`A).
This result highlights the importance of using multiple metrics of biodiversity for establishing conservation priorities.

@@

One element of biodiversity that we did not consider here is hybrids.
The Japanese fern flora includes `r n_hybrid_fern_taxa` hybrid taxa [@Ebihara2019b].
Although hybrids are often thought of as evolutionary "dead-ends", in some cases they are able to interbreed with diploids and thus contribute to diversification [@Barrington1989].
Future studies should consider the distribution of hybrids in Japan and their possible effects on biodiversity.
Methods for calculating PD would need to be modified to take into account multiple lineages contributing to one terminal taxon.

Finally, a long under-appreciated aspect of fern ecology is the role of the gametophyte.
Unlike seed plants, ferns have gametophytes that are capable of growing independently from the sporophyte.
Futhermore, the two stages of the life cycle have vastly different morphology and ecophysiology, and may even occur over partially to completely disjunct ranges [reviewed in @Pinson2016a].
Here, as in the vast majority of fern ecology studies, we consider only the sporophyte stage.
However, recent studies in Japan [@Ebihara2013; @Ebihara2019a; @Murakami2021] and elsewhere [@Nitta2017] have revealed different patterns in the distribution of gametophytes and sporophytes.
The comprehensive molecular sampling of sporophytes in Japan makes this area ideal for conducting high-throughput DNA sequencing analyses to compare patterns of biodiversity between life stages in ferns across large spatial scales.

@@

The authors thank Members of the Iwasaki Lab at the University of Tokyo and two anonymous reviewers for providing comments that improved the manuscript.
Shawn Laffan provided advice about CANAPE.
This study was supported by JSPS KAKENHI Grant Number 16H06279.

@@

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_spatial_phy.
A Docker image to run the code is available at https://hub.docker.com/r/joelnitta/japan_ferns_spatial_phy.
Data files needed to run the analysis and selected results files are available from Figshare at http://doi.org/10.6084/m9.figshare.16655263.
